Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-10-06 formally.tex
updating the section on determinism. i only describe the predicate intuitively in english, rather than giving a full description of it mathematically. i don't know if this is appropriate.

>>>>

\subsection{Determinism}
A last useful property to consider is \emph{action determinism}, i.e. that
performing an action produces a unique result. Formally, this would be
established as follows:
\begin{theorem}[Action Determinism]
  \label{thrm:actdet} All of the following hold:
  \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
  \item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and
    $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
    $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
    $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
    $\zexp' = \zexp''$ and $\htau' = \htau''$.
  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
    $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
    $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
    \zexp''$.
  \end{enumerate}
\end{theorem}
This is not a theorem of the system exactly as described so far. The reason
is quite subtle: the semantics of two construction actions,
$\aConstruct{\fasc}$ and $\aConstruct{\flam{x}}$, behave differently in the
analytic case than they do in the synthetic case -- the difference has to
do with what the generated ascription contains. Both cases can technically
``fire'' due to action subsumption, but one corresponds to an over-eager
use of subsumption: it's intended to be a rule of last resort.

The mechanization defines a predicate on action derivations that bans
exactly these over-eager uses of action subsumption, gives a mapping from
any derivation into one that matches the source and action and satisfies
the predicate, and proves determinism as stated above but only for
derivations modulo that predicate. This fix is analogous to the difference
between a chaining and focused logic in proof theory, which correspond
closely with bidirectional type checking systems, and therefore is to be
expected.\cite{Simmons11tr}
